Red Wind Quality by Justin Brown
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(gridExtra)
library(GGally)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Load_the_Data}
# Load the Data
rw <- read.csv('wineQualityReds.csv')

```

# Data Set: Red Wine Quality 

(Reference: http://www.sciencedirect.com/science/article/pii/S0167923609001377?via%3Dihub)

As a wine enthusiast, it often intrigues me the number of adjectives that wine
experts use to describe different wines.  And then for those experts to 
masterfully sort through the myrad of flavor components and arrive at a quality
rating is at the same time baffling as it is astonishing, when multiple experts
come to agree that the same wine has the same level of quality.  To my novice
tastes, I am often embarrased to admit that I can't discern between a 3 and an
8.

This data set contains 1599 independent observations across 13 variables.
Citing from the notes with this data set, the variables described are:

   1 - fixed acidity (tartaric acid - g / dm^3)
   
   2 - volatile acidity (acetic acid - g / dm^3)
   
   3 - citric acid (g / dm^3)
   
   4 - residual sugar (g / dm^3)
   
   
   5 - chlorides (sodium chloride - g / dm^3
   
   6 - free sulfur dioxide (mg / dm^3)
   
   7 - total sulfur dioxide (mg / dm^3)
   
   8 - density (g / cm^3)
   
   9 - pH
   
   10 - sulphates (potassium sulphate - g / dm3)
   
   11 - alcohol (% by volume)
   
   Output variable (based on sensory data):
   
   12 - quality (score between 0 and 10)

I am anxious to see which (if any) of these variables ultimately come together
to determine quality rating.

# Univariate Plots Section

First up is a general overview of the data.  A quick look at the first few rows 
to see what values are there to work with followed by a summary of each variable.

```{r echo=FALSE, message=FALSE, warning=FALSE, Summary_Info}
# First we will explore the content of the file using head to see the first few
# rows of data.

head(rw)

# Next I will run summary to explore the statistical information for each 
# variable.
summary(rw)

```

The following set of plots will explore each of the variables independently
to understand each a little better.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots}

p1 <- ggplot(aes(x = fixed.acidity),
  data = rw) +
  geom_histogram()
p2 <- ggplot(aes(x = volatile.acidity),
  data = rw) +
  geom_histogram()
p3 <- ggplot(aes(x = citric.acid),
  data = rw) +
  geom_histogram()
p4 <- ggplot(aes(x = residual.sugar),
  data = rw) +
  geom_histogram()
p5 <- ggplot(aes(x = chlorides),
  data = rw) +
  geom_histogram()
p6 <- ggplot(aes(x = free.sulfur.dioxide),
  data = rw) +
  geom_histogram()
p7 <- ggplot(aes(x = total.sulfur.dioxide),
  data = rw) +
  geom_histogram()
p8 <- ggplot(aes(x = density),
  data = rw) +
  geom_histogram()
p9 <- ggplot(aes(x = pH),
  data = rw) +
  geom_histogram()
p10 <- ggplot(aes(x = sulphates),
  data = rw) +
  geom_histogram()
p11 <- ggplot(aes(x = alcohol),
  data = rw) +
  geom_histogram()
p12 <- ggplot(aes(x = quality),
  data = rw) +
  geom_histogram()

grid.arrange(p1, p2, p3, p4, p5, p6, p7, p8, p9, p10, p11, p12, ncol=3)

```


# Univariate Analysis

Consistent with the notes from the authors, the median wine quality from the 
selection is 6 with a mean of 5.6.  The authors note that this is an "ordered
not balanced" data set, "(e.g. there are munch more normal wines than excellent
or poor ones)"  Also, relatively normally distributed variables appear to be
related to acidity (both fixed and volatile), density and pH.  The remaining
variables show longer tail tendencies, with some tails longer than others.
Specifically, Sugar, Chlorides, Sulphates and Sulfur Dioxide (both free and
total) are all particularly concentrated at the low end of the scale.  Finally,
Citric Acid and Alcohol content show negative, almost linear trends as one 
moves from the lower end to the higher end of the scale.

Because of the tidyness of this data, no additional clean up is required.  I am
instantly compelled to analyze whether any of the clear patterns present in the
variable data lead to corrolations in the quality ratings.  Also, of interest,
are any correlations between these chemical variables.  Perhaps they all act
independently depending on the wine maker's whim, or perhaps decisions by the
wine maker in one area of chemical make-up result in correlated attributes in
other areas represented.

# Bivariate Plots and Analysis

To begin the analysis of dependency between the variables, I will avoid the
temptation to jump to conclusions that my eyes might lead to and, instead, 
allow the data to speak for itself by looking at a 


```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=10, fig.width=10, Bivariate_Plots}

# Following is a basic ggpairs analysis to look for correlations.
ggpairs(rw)
```

Since correlation is most significant when r values are closer to +/- 1, and 
least significant when r values are closer to 0, the first points of interest
are those correlations with an absolute value closest to 1.  Also, as the 
main point of this analysis is quality prediction, seeing which variables seem
most correlated with quality is the initial thrust for this section.

The most relevant quality correlation of any variable collected is alcohol
content, having an r value of 0.476, very near the "rule of thumb" for a
significant correlation (which is r values from 0.5 to 1.0).  However, while
this correlation is not quite at 0.5, it is clear that alcohol content alone
does not determine the quality.  Let's take a closer look at this relationship.

```{r echo=FALSE, message=FALSE, warning=FALSE, Alcohol_vs_Quality}
ggplot(aes(x = alcohol, y = quality),
       data = rw) +
    geom_point(position = 'jitter', alpha = .25)
```

Interestingly, one can see that almost no wine with a rating above 6 had an 
alcohol content of less than 10% or greater than 13%, if one were to take
only this variable into account when crafting a new red wine with hopes of 
a quality rating on the higher end of the scale, this 10%-13% range seems to
be the range one would want to target.  However, as mentioned before, this
relationship alone does not determine quality as many wines in this range were
also rated 5 and 6.  However, this data does seem to indicate that wines between
12% and 13% were almost the only wines receiving a 6 or better rating, and, 
absent a few outliers, if a wine maker can determine the right balance of other
factors, almost no wine with an alcohol content greater than 13% achieved a 
quality rating less than 6.

Let's analyze three other variables that have a notable (though far from 
significant) correlation to quality, especially when compared to the 
correlations of the remaining variables. These variables are volatile.acidity,
sulphates and citric acid (in order of apparent correlation).  The r values
for these variables when compared to quality are -0.391, 0.251 and 0.226, 
respectively. Let's see these charts together.

```{r echo=FALSE, message=FALSE, warning=FALSE, Grid_of_Correlated}

# This plot will compare volatile acidity with quality
p1 <- ggplot(aes(x = volatile.acidity, y = quality),
       data = rw) +
      geom_point(position = 'jitter', alpha = .25) +
      geom_smooth()

# This plot will compare sulphates with quality
p2 <- ggplot(aes(x = sulphates, y = quality),
       data = rw) +
      geom_point(position = 'jitter', alpha = .25) +
      geom_smooth()

# This plot will compare citric acid with quality
p3 <- ggplot(aes(x = citric.acid, y = quality),
       data = rw) +
      geom_point(position = 'jitter', alpha = .25) +
      geom_smooth()

grid.arrange(p1, p2, p3, ncol=1)

```

While the correlation between citric.acid and quality is, visually, confirmed
to be weak, interestingly (with the exception of some outliers on the high end
of each scale) both volatile.acidity and sulphates appear to have some more
significant correlations in the lower to middle part of each scale.  Let's 
take another look at these charts without the outliers (defined as values above
1.0 for both). Notably, none of these outliers achieved the highest rating
of 8, so eliminating these seems to be a lower risk in determining falues
which will correlate to higher quality ratings.  In addition to eliminating the 
outliers to the right of each graph, I will zoom in on the relevant section of
data by also eliminating values for each which are lower than 0.2, as there are
very few samples affected by this limit.

```{r echo=FALSE, message=FALSE, warning=FALSE, Correlation_Zoom}

# This plot will zoom into the volatile acidity ranges between 0.2 and 1.0
p1 <- ggplot(aes(x = volatile.acidity, y = quality),
       data = rw) +
      geom_point(position = 'jitter', alpha = .25) +
      coord_cartesian(xlim = c(0.2,1.0)) +
      geom_smooth()

# This plot will zoom into the sulphate levels betwee, 0.2 and 1.0
p2 <- ggplot(aes(x = sulphates, y = quality),
       data = rw) +
      geom_point(position = 'jitter', alpha = .25) +
      coord_cartesian(xlim = c(0.2,1.0)) +
      geom_smooth()

grid.arrange(p1, p2, ncol=1)

```

By eliminating the outliers, the correlations appear much more similar to the
alcohol correlation evaluated previously.  Using this subset, I will look at
the recomputed correlations.

```{r echo=FALSE, message=FALSE, warning=FALSE, Calculate_subset_Correlations}

# In this section, we will subset the data into 2 subsets, limiting the 
# volatile.acidity and sulphates to a range of 0.2 and 1.0, respectively.
# We will then re-run correlation tests on these subsets to see if the 
# correlation is improved with the elimination of susspected outliers.

rw_subset_va <- subset(rw, volatile.acidity >= .2 & volatile.acidity <= 1.0,
                       select = c(volatile.acidity, quality))
rw_subset_sul <- subset(rw, sulphates >= .2 & sulphates <= 1.0,
                       select = c(sulphates, quality))

cor.test(x = rw_subset_va$volatile.acidity, y = rw_subset_va$quality)
cor.test(x = rw_subset_sul$sulphates, y = rw_subset_sul$quality)

```

Although the outliers visually appeared they may be affecting both correlations,
Pearson's test shows that eliminating the outliers only improves one of the two
correlations.  And it is a fairly significant increase from 0.251 to 0.394 when
analyzing sulphates.

The relationship between citric.acid and volatile.acidity (r = -0.552) initially
escaped scrutiny.  As both are a measure of acidity, is would not be unexpected
to see correlation.  However, what was interesting is that these variables are
negatively correlated, not positively as one might expect based on title alone.
Further, since both of these variables raised interest in terms of their 
correlation to quality, at least a surface level review is warranted.  After
all, if one of these measures would be easier for a wine maker to control than
the other, assuming their high correlation to each other and relatively higher
correlation to quality as compared to other variables, this information may be
useful in driving the proper make-up of a high quality wine.

```{r echo=FALSE, message=FALSE, warning=FALSE, Citric_vs_Volatile}

# This plot will explore the relationship between citric.acid and
# volatile.acidity.

ggplot(aes(x = citric.acid, y = volatile.acidity),
       data = rw) +
  geom_point(position = 'jitter', alpha = .25) +
  geom_smooth()
```

Based on this relationship and the corresponding relationships to quality, if
one could more easily control for citric.acid content, higher levels of citric
acid will drive lower levels of volatile acidity, both of which appear more
strongly correlated with wines that rate higher on the quality scale.  The 
inverse is true if a wine maker could more easily control volatile acidity. 
That is lower levels of volatile acidity correlate with higher levels of 
citric acid, which in turn correlates to wines that tend to rate higher on the
quality scale.

Correlations between the variables other than quality do exist, however, due
to the relativity of the quality measured, these correlations do not seem 
interesting initially and will fall outside of this analysis.  They may,
however, be worth a further look in a more extensive review.  Following is 
a list of some of these correlations with a brief mention of the rationale for 
their exclusion in this review:
* ph and fixed.acidity (r = -0.683) - While this is the highest correlation
observed between any two variables in the data set, the correlation is not
illogical.  pH is a measure of acidity (or alkalinity) and lower pH values
are indicative of higher levels of acidity, therefore a relatively high negative
correlation would be expected.
* total.sulfur.dioxide and free.sulfur.dioxide (r = 0.668) - Because both of 
these data points relate to sulfur dioxide, a strong positive correlation is 
not illogical.
* density and fixed.acidity (r = 0.668) - This relationship is more interesting,
however, as neither density nor fixed.acidity seem to have a high correlation
to the quality (which is the focus of this analysis), this relationship will
not be explored further.


# Multivariate Plots Section

The following plots investigate more complex relationships between the variables
highlighted above.

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots}

# Analyze whether citric.acid levels have different quality tendancies
# across the range of sulphate levels.

ggplot(aes(x = sulphates, y = quality, color = citric.acid),
       data = rw) +
      geom_point(position = 'jitter', alpha = 0.1, size = 10) +
      coord_cartesian(xlim = c(0.2,1.0)) +
      scale_colour_gradient(low = "black", high = "red") +
      ggtitle('Sulphates and Citric Acid to Quality')

# Analyze whether sulphate levels have different quality tendancies
# across the range of citric.acid levels.

ggplot(aes(x = citric.acid, y = quality, color = sulphates),
       data = rw) +
      geom_point(position = 'jitter', alpha = 0.1, size = 10) +
      coord_cartesian(xlim = c(0,1.0)) +
      scale_colour_gradient(low = "black", high = "red") +
      ggtitle('Citric Acid and Sulphates to Quality')

# Analyze whether volatile.acidity levels have different quality tendancies
# across the range of citric.acid levels.

ggplot(aes(x = citric.acid, y = quality, color = volatile.acidity),
       data = rw) +
      geom_point(position = 'jitter', alpha = 0.1, size = 10) +
      coord_cartesian(xlim = c(0,1.0)) +
      scale_colour_gradient(low = "black", high = "red") +
      ggtitle('Citric Acid and Volatile Acidity to Quality')

# Analyze whether citric.acid levels have different quality tendancies
# across the range of alcohol levels.

ggplot(aes(x = alcohol, y = quality, color = citric.acid),
       data = rw) +
      geom_point(position = 'jitter', alpha = 0.5, size = 10) +
      coord_cartesian(xlim = c(9,14)) +
      scale_colour_gradient(low = "black", high = "red") +
      ggtitle('Citric Acid and Volatile Acidity to Quality')

```

# Multivariate Analysis

Relating these three variables which appear to most impact quality, I did not
notice any combinations that appeared to strengthen the correlation between
the variables and quality.  Using color to indicate whether one variable tended
to more stronly impact quality when combined with another variable, there was
no clear example where, for fixed values of the x variable, the color variable
tended toward one end or the other of the quality scale.

Maybe most intersting is that the negative correlation between citric acid and
volatile acidity and their combined effect on quality was not evidenced by the 
multivariate plots.

------

# Final Plots and Summary

In concluding this exploration, the following three plots appear to show
the strongest relationships between quality and other variables.

### Plot One

```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_One}

# Revisiting the alcohol vs quality relationship using jitter, alpha and size
# to highlight areas of concentration.

ggplot(aes(x = alcohol, y = quality),
       data = rw) +
    geom_point(position = 'jitter', alpha = .5, size = 10) +
    scale_x_continuous(limits = c(8,15), breaks = seq(0,15,1)) +
    labs(title = 'A Comparison of Alcohol Content to Quality',
         x= 'Alcohol Content',
         y = 'Quality')
```

### Description One

For my initial summary plot, I go all the way back to one of the first plots in
this analysis.  At first, this plot did not seem so relevant in light of the 
apparent lack of a strong correlation and the evidince of stronger correlations
between quality and other variables.  However, in the end, within this sample,
controlling for alcohol content actually apears to be one way to achieve a 
higher possibility of a good rating.

Specifically, across 1599 observations, all oservations between around 13.25%
and 14% alcohol achieved ratings of 6 or higher.  This seems to be a sweet spot
at which, if controlled, can provide a strong opportunity for quality at a
a higher level.  Further very few wines had an alcohol content of lowere than 9%
and of those, none achieved a rating above 6.  This is territory to avoid.

For commodity wines, alcohol content between 9 and 13 will put you in the 
middle of the pack.

### Plot Two

```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Two}

# Revisiting the volatile.acidity vs quality relationship using jitter, alpha
# and size to highlight areas of concentration.

ggplot(aes(x = volatile.acidity, y = quality),
       data = rw) +
      geom_point(position = 'jitter', alpha = .25, size = 10) +
      scale_x_continuous(breaks = seq(0,1.6,0.2)) +
      labs(title = 'A Comparison of Volatile Acidity to Quality',
         x= 'Volatile Acidity',
         y = 'Quality')
```

### Description Two

This is one of the more obvious ways to improve the chances of a higher quality
rating, or more specifically, how to avoid a mediocre or low rating.  There is
a very high concentration of 7 and 8 quality ratings between 0.2 and 0.4.  
While evidence of lower ratings also exist in this range, the likelihood of 
achieving a 7+ rating is significantly reduced above a volitile acidity level
of 0.45.

### Plot Three
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Three}

# Revisiting the sulphates vs quality relationship using jitter, alpha and size
# to highlight areas of concentration.

ggplot(aes(x = sulphates, y = quality),
       data = rw) +
      geom_point(position = 'jitter', alpha = .5, size = 10) +
      scale_x_continuous(breaks = seq(0,1.6,0.2)) +
      coord_cartesian(xlim = c(0.2,1.0)) +
      labs(title = 'A Comparison of Sulphates to Quality',
         x= 'Sulphates',
         y = 'Quality')
```

### Description Three

Lastly, the comparison between sulphates and quality identifies a target range
to aim for in an attempt to improve your rating.  No 8 rated wines had a 
sulphate level less than 0.6 and only few 7 rated wines had a level below 0.5.

Once again, controlling sulphates to greater than 0.5 doesn't guarantee a 7+
rating... there are many 5 and 6 rated wines in this range as well.  But the 
tasters appear to favor sulphates between 0.5 and 0.9 in the wines that recieved
the more favorable ratings.

------

# Reflection

Reflecting on the exploratory analysis above, I am left to wonder if the rating
a wine recieves might be as much a factor of the taster as any combination of
qualities.  That said, there do seem to be a few factors (volatile acidity,
sulphates and alcohol content) for which, if one places a new wine in favorable
ranges, tend to gain favorale quality ratings as a result.

In order to better understand the impact of the individual wine judge, it would
be good to have a panel of ratings for each wine, with the same panel of judges
performing all observations.  One could then identify a balanced observation
controlling for different taste preferences, as well as identify if certain
individuals tend to have unique and consistent taste preferences across a range
of observations.

In the end, I know what I would target for a wine creation that I hoped would
appeal to wine tasters:

* Alcohol between 13% and 14%

* Volatile Acidity between 0.2 and 0.4

* Sulphates between 0.7 and 0.9
